Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation

### 2022-04-18 ms/manuscript.Rmd
Update MS
according to editor's comments

---
title: "Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions:
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
output:
  bookdown::word_document2:
    pandoc_args: [ "--csl", "american-journal-of-botany.csl"]
    toc: no
    number_sections: no
    reference_docx: american-journal-of-botany.docx
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: no
    fig_caption: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "american-journal-of-botany.csl"]
params:
  doc_type: doc
editor_options:
  chunk_output_type: console
---

@@

Joel H. Nitta^1,5^, Brent D. Mishler^2^, Wataru Iwasaki^1,3^, and Atsushi Ebihara^4^

^1^Department of Biological Sciences, Graduate School of Science, The University of Tokyo, Tokyo, Japan

@@

^3^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan

@@

^5^Author for correspondence: joelnitta@gmail.com

```{r author-page-doc, results = "asis", echo = FALSE, eval = params$doc_type == "doc"}
# Author page, version for submission to journal
cat('
Manuscript received _______; revision accepted _______.

Short running title: Spatial phylogenetics of Japanese ferns
')
```

@@

**Premise:** Biodiversity is often only measured with species richness.
However, this ignores evolutionary history and is not sufficient for making conservation decisions.
Here, we characterize multiple facets and drivers of biodiversity to understand how these relate to bioregions and conservation status in the ferns of Japan.

**Methods:** We compiled a community dataset of `r nrow(comm_ferns) %>% scales::number(big.mark = ",")` 20 km × 20 km grid-cells including `r ncol(comm_ferns)` taxa based on > `r occ_sum$filtered$n_specimens$value %>% plyr::round_any(1000, f = floor) %>% number(big.mark = ",")` specimen records.
We combined this with a phylogeny and functional traits to analyze taxonomic, phylogenetic, and functional diversity, and modeled biodiversity metrics in response to environmental factors and reproductive mode.
Hierarchical clustering was used to delimit bioregions.
Conservation status and threats were assessed by comparing the overlap of significantly diverse grid-cells with conservation zones and range maps of native Japanese deer.

**Results:** Taxonomic richness was highest at mid-latitudes.
Phylogenetic and functional diversity and phylogenetic endemism were highest in small southern islands.
Relative phylogenetic and functional diversity were high at high and low latitudes, and low at mid-latitudes.
Grid-cells were grouped into `r maj_cluster_count$phylo_cluster$n %>% english2` (phylogenetic) or `r maj_cluster_count$taxonomic_cluster$n %>% english2` (taxonomic) major bioregions.
Temperature and apomixis were identified as drivers of biodiversity patterns.
Conservation status was generally high for grid-cells with significantly high biodiversity, but the threat due to herbivory by deer was greater for taxonomic richness than other metrics.

@@

Characterizing the spatial distribution of biodiversity is a major goal of evolutionary biology with two equally important aspects: to understand the processes generating it, and to conserve it.
Until recently, the vast majority of studies seeking to understand the spatial distribution of biodiversity have focused on species richness.
Hotspots, or areas of exceptionally high species richness or endemism, have received particular attention, resulting in a widely recognized set of 36 terrestrial hotspots [@Myers2000; @Noss2016] and motivating conservation strategies to preserve them [@Margules2000; @Brooks2006].
However, species richness alone cannot provide a complete picture of biodiversity [@Miller2018].
All organisms are related to a greater or lesser degree by descent from a common ancestor, and these evolutionary relationships must be taken into account to obtain a full understanding of the biodiversity present in an area.

Presumably, the main reason phylogeny has not been taken into account more prominently in biodiversity studies is because the necessary data (DNA sequences and georeferenced occurrence records) and analytic tools have only become available relatively recently [@Soltis2016a; @Folk2021].
These datasets and tools now make it possible to analyze other dimensions of biodiversity, such as phylogenetic diversity [@Faith1992] and phylogenetic endemism [@Rosauer2009].
Better ways to characterize biological regions ("bioregions"; `r ie`, areas defined by their taxonomic composition or evolutionary history) are also available [@Laffan2016; @White2019; @Daru2020a], rather than relying on ad-hoc characterizations.
Furthermore, incorporating these two frameworks---the categorization of areas into phyloregions with analysis of over/under dispersion of biodiversity---can provide powerful insights into the processes structuring biodiversity and suggest conservation priorities.

However, a comprehensive understanding of the relationships between richness and other metrics requires densely sampled data, and such datasets are rare on the regional (country) scale.
The ferns of Japan are excellent model system because they have been the target of intense botanical interest for several decades and are densely sampled [reviewed in @Ebihara2019b].
The ferns of Japan include `r total_fern_taxa` native, non-hybrid taxa (including species and varieties) and hundreds of hybrids [nothotaxa\; @Ebihara2019b].
The availability of detailed distribution data [distribution maps at the ca. 10 km scale for nearly all species\; @Ebihara2016b; @Ebihara2017], trait data [multiple quantitative and qualitative traits compiled for identification of nearly all species\; @Ebihara2019b], and DNA sequences for > 97% of species [@Ebihara2010; @Ebihara2019b\; all coverage statistics exclude nothotaxa] make the ferns of Japan an ideal system for investigating the relationships between, and drivers of, multiple dimensions of biodiversity.

One particularly valuable characteristic of the Japanese fern flora is the availability of data on reproductive mode, which is known for `r repro_count$known$total` native fern taxa excluding hybrids [`r repro_count$known$percent`\; @Ebihara2019b].
Reproductive mode is likely to affect population-level genetic diversity [@Bengtsson2003], and thereby higher-level biodiversity [@Krueger-Hadfield2019].
In ferns, clonally reproducing apomictic species are often polyploid hybrids that share identical plastid genotypes among taxa and within populations [@Grusz2009; @Chao2012a; @Hori2014; @Hori2019].
Therefore, we expect high prevalence of apomictic species within a community to decrease phylogenetic diversity.

@@

Here we leverage this exceptionally rich dataset to analyze the geographic distribution of biodiversity in detail.
We ask the following questions in our study system, the ferns of Japan: 1. How is biodiversity distributed? 2. How is biodiversity structured? 3. What environmental and biological factors influence biodiversity? 4. How well is biodiversity protected?

@@

We used a list of native, non-hybrid fern specimens deposited at herbaria in Japan to quantify occurrences [@Ebihara2016b; @Ebihara2017; @Ebihara2019b].
We chose to use this over other sources such as the Global Biodiversity Information Facility (GBIF) because of its high quality, which obviates many cleaning steps otherwise needed when working with large, publicly available datasets.
Furthermore, the overall sampling completeness of this dataset is also high (`r s_figure("sampling-curve")` in Appendix S1; see Supplemental Data with this article).
The original list comprises `r occ_sum$unfiltered$n_specimens$value %>% scales::number(big.mark = ",")` specimens representing `r occ_sum$unfiltered$n_taxa$value` terminal taxa at the rank of species, subspecies, form, or variety, excluding non-native taxa and nothotaxa.
We treated infraspecific taxa as distinct species during analysis and hereafter use "taxon" and "species" interchangeably to refer to entities at this rank, unless otherwise indicated.
All names are standardized to a common taxonomy, the Japan Green List v. 1.01 (http://www.rdplants.org/gl/).

@@

We further cleaned the list prior to analysis by filtering out any occurrences not within the second-degree mesh and removing duplicate collections (`r occ_sum$filtered$n_specimens$value %>% scales::number(big.mark = ",")` specimens, `r occ_sum$filtered$n_taxa$value` taxa after filtering).
Given the high quality of our occurrence data and the fact that automated occurrence cleaning algorithms [`r eg`, CoordinateCleaner\; @Zizka2019] have the potential to erroneously exclude true occurrence points [`r ie`, false positives\; @Zizka2020], we chose not to apply additional automated cleaning steps to our data as is often done with occurrence records obtained from GBIF [`r eg`, @Rice2019a; @Suissa2021].

A necessary step in any analysis of biodiversity is to set the grain size used to accurately define communities (co-occurring species).
Here, one must consider that smaller grain size is needed to detect environmental effects, while larger grain size is needed to ensure adequate species sampling.
We created sets of 10 km, 20 km, 30 km, and 40 km grid-cells covering Japan using a Mollweide equal-area projection (size refers to the length of the side of each square grid-cell).
At each grain size, species occurrences were converted to a presence-absence community matrix (a species was considered present if at least one specimen was recorded in that grid-cell).
We calculated sampling redundancy [1 - richness/number of specimens\; @Garcillan2003] to quantify adequacy of sampling.
Preliminary analysis indicated that 20 km grid-cells are optimal for our dataset: there is a sudden improvement in redundancy values from 10 km to 20 km, but much less improvement as grain size is increased beyond 20 km (`r s_figure("grain-size")`).
Although the grid-cells are defined with equal area, actual land area of each grid-cell varies due to coastlines.
Strictly filtering out all grid-cells with less than complete land area would result in a large loss of data, as Japan has many coastlines and small islands.
Therefore, we instead filtered the grid-cells by sampling completeness, removing any grid-cells with redundancy < 0.1; this dataset was used for all subsequent analyses.
The R package `r pack("sf")` was used for all GIS analyses [@Pebesma2018].

@@

We used traits originally compiled for identification of ferns and lycophytes of Japan [@Ebihara2016b; @Ebihara2017], which were formatted to be used with Lucid software (https://www.lucidcentral.org/).
Continuous traits were measured on 10 randomly selected specimens per species, then four values per species were obtained from these measurements following Lucid format: outside (`r ie`, outside the typical range) minimum, typical minimum, typical maximum, and outside maximum.
For species with dimorphic fertile (`r ie`, spore-bearing) and sterile fronds, fertile and sterile fronds were measured separately.
We took the mean of the typical minimum and maximum values to use as the species mean value.
Qualitative traits were scored by observing voucher specimens.
All qualitative traits were scored in binary format; `r eg`, a single trait with three states was formatted as three binary traits.
From the original trait list, we selected only traits with putative ecological function (`r table("traits-used")`) and excluded any traits that had Pearson correlation coefficient > 0.6 or fewer than three observations of a given trait state.

@@

We used data compiled by @Ebihara2019b, which classifies each fern taxon as sexual, apomictic, mixed (both sexual and apomictic modes known from the same taxon), or unknown.
We calculated the percentage of apomictic taxa in each grid-cell as the sum of apomictic and mixed taxa divided by total number of taxa (`r s_figure("envir")`B).

@@

We downloaded climate data at the 2.5 minute scale from the WorldClim database using the `r func("getData")` function in the R package `r pack("raster")` [@R-raster], and extracted four variables related to our research questions as follows: mean annual temperature (bio1; hereafter "temperature"), temperature seasonality (bio4), annual precipitation (bio12; hereafter "precipitation"), and precipitation seasonality (bio15) (`r s_figure("envir")`).
We identified the intersection of the climate data at the 2.5 minute scale with the 20 km grid-cells, then calculated the mean of each climatic variable for each grid-cell.
We calculated area as rolling mean area (km^2^) across 100 km latitudinal bands.
This approach reflects the idea that the amount of area with a given climate regime enclosing each grid-cell may influence species richness of that cell [@Fahrig2013], and that using raw island areas would not be appropriate in this context as there are just four very large islands with many much smaller ones located close by (`r figure("japan-map")`).

@@

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]; these sequences originate from samples that were identified using the same taxonomic system as the occurrence data [@Ebihara2010; @Ebihara2019b], so we used these preferentially over other data available on GenBank that might include misidentifications or different taxon concepts.
However, using this dataset alone for phylogenetic analysis suffers two drawbacks: it is not densely sampled enough to resolve some internal nodes, and it lacks many lineages that have fossils available for molecular dating.
Furthermore, use of community phylogenies generated by only by sampling the species present in the local community has been shown to produce spurious results in simulation studies [@Park2018].
Because one goal of the study is to understand the distribution of paleo- vs. neo-endemic areas (defined in units of time), we require an ultrametric tree.
Therefore, to obtain a robust, ultrametric tree, we combined the *rbcL* data of @Ebihara2019b (`r n_japan_fern_taxa` taxa) with a globally sampled plastid dataset including all fern sequences on GenBank for four widely-sequenced plastid genes (*atpA*, *atpB*, *rbcL*, and *rps4*; `r n_ftol_fern_taxa` taxa) and `r n_plastome_genes` other coding, single-copy plastid genes extracted from `r n_plastome_fern_taxa` complete fern plastomes (FTOL v0.0.1; Nitta et al., in prep.; https://github.com/fernphy/ftol/) (`r ape::Ntip(plastome_tree) %>% scales::number(big.mark = ",")` OTUs total, including outgroup taxa).
This gene sampling is comparable to a recent global fern phylogeny that resolved relationships across ca. 4,000 taxa using six plastid markers [@Testo2016a], and the addition of plastome data can be expected to increase support along the backbone.

We conducted maximum-likelihood phylogenetic analysis with `r software("IQ-TREE")` v1.6.12 [@Nguyen2015] under the GTR+I+G model (all sites concatenated into a single data matrix), and assessed node support with 1,000 ultrafast bootstrap replicates [@Minh2013].
Molecular dating was conducted with `r software("treePL")` v1.0 [@Smith2012] using `r n_fossil_points` fossil calibration points after @Testo2016a, with the exception of treating fossil *Kuylisporites* as belonging to crown *Cyathea*, not *Cyathea* + *Alsophila* [@Loiseau2020].
The root of the tree (most recent common ancestor of all land plants) was fixed to 475 mya [@Testo2016a].
We trimmed the resulting ultrametric tree to only Japanese taxa.
`r software("treePL")` requires branch lengths to be > 0, and it sets extremely small branches to an arbitrary minimum positive length.
This may artificially add branch length between taxa with identical DNA sequences, which occur multiple times in the Japanese fern *rbcL* dataset.
We converted any clades in the dated tree consisting of identical sequences to polytomies with the node at 0 mya (`r ie`, the present).
This tree was used for all subsequent analysis unless indicated otherwise.

@@

We tested for phylogenetic signal in continuous traits using two alternative metrics, Blomberg's *K* [@Blomberg2003] and Pagel's lambda [@Pagel1999].
*K* describes the ratio between the observed variance in a trait vs. the amount of variance expected under Brownian Motion (BM); lambda is a scaling parameter between 0 and 1 that transforms the phylogeny such that trait values fit those most closely expected under BM.
Both lambda and *K* = 1 under BM; relatively lower values indicate less signal (overdispersion) and higher values indicate more signal (clustering).
We measured *K* and lambda using the `r func("phylosig")` function in the R package `r pack("phytools")` [@Revell2012], which also calculates significance based on a randomization test for *K* and a likelihood ratio test for lambda.
We tested for phylogenetic signal in qualitative (coded as binary) traits using Fritz and Purvis' *D* [@Fritz2010] with the `r func("phylo.d")` function in the R package `r pack("caper")` [@Orme2018], which also assesses significance by comparing the observed value to a null distribution of random values and a simulated distribution of values expected under BM.
Values of *D* range from 0 under BM to 1 under random evolution, but can exceed this range in cases of extreme clumping or overdispersion, respectively.
These methods for phylogenetic signal cannot use a tree with zero-length polytomies, so we used the original dated tree from treePL.

@@

We measured taxonomic diversity using taxonomic richness (number of taxa in each community).

@@

We measured functional diversity (FD) following the method of @Petchey2002.
We first calculated functional distances using Gower's method [@Gower1971] on the trait data with the `r func("gowdis")` function in the R package `r pack("FD")` [@Laliberte2010].
We weighted continuous and qualitative traits equally, weighted each binary component trait of each qualitative trait equally, and log-transformed and scaled continuous traits prior to calculating distances.
We also tested alternative weighting schemes, but these produced very similar results (data not shown).
We then used the distances to build a trait dendrogram using the `r func("hclust")` function in the R package `r pack("stats")` [@RCoreTeam2021], and used the dendogram in place of a phylogenetic tree to calculate PD and RPD as described above.
These functional analogs of PD and RPD are hereafter referred to as FD (functional diversity) and RFD (relative functional diversity).

@@

Observed values of PD, RPD, FD, RFD, and PE are known to be related to species richness: as more taxa are drawn at random without replacement from a given set, it becomes increasingly unlikely to draw a taxon that is distantly related (or functionally dissimilar) to the others.
Therefore, we determined whether the observed metrics are more or less diverse than random for a given number of taxa by conducting a randomization test.
We generated 999 random communities using the "curveball" algorithm of @Strona2014, which conserves richness per site and species occurrence frequencies while randomizing species identities, then compared the observed value to the distribution of random values.
Statistical significance was assessed using a one-tailed test for PE or a two-tailed test for other metrics with alpha = 0.5; observed values in the extreme 5% of the null distribution were considered significant.
We also calculated the standard effect size (SES) of each biodiversity metric.
SES measures how extreme the observed value (*obs*) is relative to the null distribution (*null*):

$SES = \frac{obs - mean(null)}{sd(null)}$

We tested whether centers of endemism have an over-representation of old (paleoendemic) or new (neoendemic) lineages using categorical analysis of neo- and paleo-endemism (CANAPE), which involves measuring PE with an alternate tree where all branch lengths are equal, then comparing the ratio between raw PE and alternative PE (RPE) [@Mishler2014].
Low RPE indicates short range-restricted branches (neoendemism) and high RPE indicates long range-restricted branches (paleoendemism).
Since all of these measures are affected by species richness, categorization of endemism is done using the rank of each observed variable relative to those calculated for the null distribution. All measures of biodiversity and CANAPE were carried out using the R package `r pack("canaper")` [@Nitta2022].

>>>>

---
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions:
  - BoldFont=Roboto-Bold.ttf
  - ItalicFont=Roboto-Italic.ttf
  - BoldItalicFont=Roboto-BoldItalic.ttf
output:
  bookdown::word_document2:
    pandoc_args: [ "--csl", "american-journal-of-botany.csl"]
    toc: no
    number_sections: no
    reference_docx: american-journal-of-botany.docx
  bookdown::pdf_document2:
    citation_package: default
    toc: no
    number_sections: no
    fig_caption: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "american-journal-of-botany.csl"]
params:
  doc_type: doc
editor_options:
  chunk_output_type: console
---

@@

RESEARCH ARTICLE

Short Title: Spatial phylogenetics of Japanese ferns

__Spatial phylogenetics of Japanese ferns: patterns, processes, and implications for conservation__

Joel H. Nitta^1^, Brent D. Mishler^2^, Wataru Iwasaki^1^, and Atsushi Ebihara^3^

```{r author-page-doc, results = "asis", echo = FALSE, eval = params$doc_type == "doc"}
# Author page, version for submission to journal
cat('
Manuscript received 22 September 2021; revision accepted 11 March 2022
')
```

^1^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan

@@

@@

__Correspondence__

Joel H. Nitta, #601 Iwasaki Lab, Bioscience Bldg, The University of Tokyo, 5-1-5 Kashiwanoha, Kashiwa, Chiba 277-0882, Japan

Email: joelnitta@gmail.com

ORCID ID: 0000-0003-4719-7472

```{r author-page-doc-2, results = "asis", echo = FALSE, eval = params$doc_type == "doc"}
# Author page, version for submission to journal
cat('
__Citation__: Nitta, J. H., B. D. Mishler, W. Iwasaki, and A. Ebihara. 2022. Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation. *American Journal of Botany* 109(5): XXXX

__DOI: XXXX__
')
```

@@

**Premise:** Biodiversity is often only measured with species richness; however, this measure ignores evolutionary history and is not sufficient for making conservation decisions.
Here, we characterize multiple facets and drivers of biodiversity to understand how these relate to bioregions and conservation status in the ferns of Japan.

**Methods:** We compiled a community data set of `r nrow(comm_ferns) %>% scales::number(big.mark = ",")` grid cells (20 km × 20 km each) including `r ncol(comm_ferns)` taxa based on > `r occ_sum$filtered$n_specimens$value %>% plyr::round_any(1000, f = floor) %>% number(big.mark = ",")` specimen records.
We combined the community data with a phylogeny and functional traits to analyze taxonomic, phylogenetic, and functional diversity and modeled biodiversity metrics in response to environmental factors and reproductive mode.
Hierarchical clustering was used to delimit bioregions.
Conservation status and threats were assessed by comparing the overlap of significantly diverse grid cells with conservation zones and range maps of native Japanese deer.

**Results:** Taxonomic richness was highest at mid latitudes.
Phylogenetic and functional diversity and phylogenetic endemism were highest in small southern islands.
Relative phylogenetic and functional diversity were high at high and low latitudes, and low at mid latitudes.
Grid cells were grouped into `r maj_cluster_count$phylo_cluster$n %>% english2` (phylogenetic) or `r maj_cluster_count$taxonomic_cluster$n %>% english2` (taxonomic) major bioregions.
Temperature and apomixis were identified as drivers of biodiversity patterns.
Conservation status was generally high for grid cells with significantly high biodiversity, but the threat due to herbivory by deer was greater for taxonomic richness than other metrics.

@@

Characterizing the spatial distribution of biodiversity is a major goal of evolutionary biology with two equally important aspects: to understand the processes generating it and to conserve it.
Until recently, the vast majority of studies seeking to understand the spatial distribution of biodiversity have focused on species richness.
Hotspots, or areas of exceptionally high species richness or endemism, have received particular attention, resulting in a widely recognized set of 36 terrestrial hotspots [@Myers2000; @Noss2016] and motivating conservation strategies to preserve them [@Margules2000; @Brooks2006].
However, species richness alone cannot provide a complete picture of biodiversity [@Miller2018].
All organisms are related to a greater or lesser degree by descent from a common ancestor, and these evolutionary relationships must be taken into account to obtain a full understanding of the biodiversity present in an area.

Presumably, phylogeny has not been taken into account more prominently in biodiversity studies mainly because the necessary data (DNA sequences and georeferenced occurrence records) and analytic tools have only become available relatively recently [@Soltis2016a; @Folk2021].
These datasets and tools now make it possible to analyze other dimensions of biodiversity such as phylogenetic diversity [@Faith1992] and phylogenetic endemism [@Rosauer2009].
Better ways to characterize biological regions ("bioregions"; `r ie`, areas defined by their taxonomic composition or evolutionary history) are also available [@Laffan2016; @White2019; @Daru2020a], rather than relying on ad hoc characterizations.
Furthermore, incorporating these two frameworks---the categorization of areas into phyloregions with analysis of over/under dispersion of biodiversity---can provide powerful insights into the processes structuring biodiversity and suggest conservation priorities.

A comprehensive understanding of the relationships between richness and other metrics requires densely sampled data, but such datasets are rare on the regional (country) scale.
The ferns of Japan are excellent model system because they have been the target of intense botanical interest for several decades and are densely sampled [reviewed by @Ebihara2019b].
The ferns of Japan include `r total_fern_taxa` native, nonhybrid taxa (including species and varieties) and hundreds of hybrids [nothotaxa\; @Ebihara2019b].
The availability of detailed distribution data [distribution maps at the ca. 10 km scale for nearly all species\; @Ebihara2016b; @Ebihara2017], trait data [multiple quantitative and qualitative traits compiled for identification of nearly all species\; @Ebihara2019b], and DNA sequences for > 97% of species [@Ebihara2010; @Ebihara2019b\; all coverage statistics exclude nothotaxa] make the ferns of Japan an ideal system for investigating the relationships between, and drivers of, multiple dimensions of biodiversity.

One particularly valuable characteristic of the Japanese fern flora is the availability of data on reproductive mode, which is known for `r repro_count$known$total` native fern taxa excluding hybrids [`r repro_count$known$percent`\; @Ebihara2019b].
Reproductive mode is likely to affect population-level genetic diversity [@Bengtsson2003] and thereby higher-level biodiversity [@Krueger-Hadfield2019].
In ferns, clonally reproducing apomictic species are often polyploid hybrids that share identical plastid genotypes among taxa and within populations [@Grusz2009; @Chao2012a; @Hori2014; @Hori2019].
Therefore, we expect high prevalence of apomictic species within a community to decrease phylogenetic diversity.

@@

Here, we leveraged this exceptionally rich data set to analyze the geographic distribution of biodiversity in detail.
We asked the following questions in our study system, the ferns of Japan: 1. How is biodiversity distributed? 2. How is biodiversity structured? 3. What environmental and biological factors influence biodiversity? 4. How well is biodiversity protected?

@@

We used a list of native, nonhybrid fern specimens deposited at herbaria in Japan to quantify occurrences [@Ebihara2016b; @Ebihara2017; @Ebihara2019b].
We chose to use this over other sources such as the Global Biodiversity Information Facility (GBIF) because of its high quality, which obviates many cleaning steps otherwise needed when working with large, publicly available datasets.
Furthermore, the overall sampling completeness of this dataset is also high (`r s_figure("sampling-curve")` in Appendix S1; see Supplemental Data with this article).
The original list comprises `r occ_sum$unfiltered$n_specimens$value %>% scales::number(big.mark = ",")` specimens representing `r occ_sum$unfiltered$n_taxa$value` terminal taxa at the rank of species, subspecies, form, or variety, excluding non-native taxa and nothotaxa.
We treated infraspecific taxa as distinct species during analysis and hereafter use "taxon" and "species" interchangeably to refer to entities at this rank, unless otherwise indicated.
All names are standardized to a common taxonomy, the Japan Green List v. 1.01 [@Ebihara2019c].

@@

We further cleaned the list before analysis by filtering out any occurrences not within the second-degree mesh and removing duplicate collections (`r occ_sum$filtered$n_specimens$value %>% scales::number(big.mark = ",")` specimens, `r occ_sum$filtered$n_taxa$value` taxa after filtering).
Given the high quality of our occurrence data and the fact that automated occurrence cleaning algorithms [`r eg`, CoordinateCleaner\; @Zizka2019] have the potential to erroneously exclude true occurrence points [`r ie`, false positives\; @Zizka2020], we chose not to apply additional automated cleaning steps to our data as is often done with occurrence records obtained from GBIF [`r eg`, @Rice2019a; @Suissa2021].

A necessary step in any analysis of biodiversity is to set the grain size used to accurately define communities (co-occurring species).
Here, one must consider that smaller grain size is needed to detect environmental effects, while larger grain size is needed to ensure adequate species sampling.
We created sets of 10-km, 20-km, 30-km, and 40-km grid cells covering Japan using a Mollweide equal-area projection (size refers to the length of the side of each square grid cell).
At each grain size, species occurrences were converted to a presence–absence community matrix (a species was considered present if at least one specimen was recorded in that grid cell).
We calculated sampling redundancy [1 - Richness/Number of specimens\; @Garcillan2003] to quantify adequacy of sampling.
Preliminary analysis indicated that 20-km grid cells are optimal for our data set; redundancy values greatly increased from 10 to 20 km, but much less improvement as grain size is increased beyond 20 km (`r s_figure("grain-size")`).
Although the grid cells are defined with equal area, actual land area of each grid cell varies due to coastlines.
Strictly filtering out all grid cells with less than complete land area would result in a large loss of data, as Japan has many coastlines and small islands.
Therefore, we instead filtered the grid cells by sampling completeness, removing any grid cells with redundancy < 0.1; this dataset was used for all subsequent analyses.
The R package `r pack("sf")` was used for all GIS analyses [@Pebesma2018].

@@

We used traits originally compiled for identification of ferns and lycophytes of Japan [@Ebihara2016b; @Ebihara2017], which were formatted to be used with Lucid software (https://www.lucidcentral.org/).
Continuous traits were measured on 10 randomly selected specimens per species, then four values per species were obtained from these measurements following the Lucid format: outside (`r ie`, outside the typical range) minimum, typical minimum, typical maximum, and outside maximum.
For species with dimorphic fertile (`r ie`, spore-bearing) and sterile fronds, fertile and sterile fronds were measured separately.
We took the mean of the typical minimum and maximum values to use as the species mean value.
Qualitative traits were scored by observing voucher specimens.
All qualitative traits were scored in binary format; `r eg`, a single trait with three states was formatted as three binary traits.
From the original trait list, we selected only traits with putative ecological function (`r table("traits-used")`) and excluded any traits that had Pearson correlation coefficient > 0.6 or fewer than three observations of a given trait state.

@@

We used data compiled by @Ebihara2019b, which classifies each fern taxon as sexual, apomictic, mixed (both sexual and apomictic modes known from the same taxon), or unknown.
We calculated the percentage of apomictic taxa in each grid cell as the sum of apomictic and mixed taxa divided by total number of taxa (`r s_figure("envir")`B).

@@

We downloaded climate data at the 2.5 minute scale from the WorldClim database using the `r func("getData")` function in the R package `r pack("raster")` [@R-raster], and extracted four variables related to our research questions as follows: mean annual temperature (bio1; hereafter "temperature"), temperature seasonality (bio4), annual precipitation (bio12; hereafter "precipitation"), and precipitation seasonality (bio15) (`r s_figure("envir")`).
We identified the intersection of the climate data at the 2.5-minute scale with the 20-km grid cells, then calculated the mean of each climatic variable for each grid cell.
We calculated area as rolling mean area (km^2^) across 100-km latitudinal bands.
This approach reflects the idea that the amount of area with a given climate regime enclosing each grid cell may influence species richness of that cell [@Fahrig2013], and that using raw island areas would not be appropriate in this context because there are just four very large islands with many much smaller ones located close by (`r figure("japan-map")`).

@@

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]; these sequences originate from samples that were identified using the same taxonomic system as the occurrence data [@Ebihara2010; @Ebihara2019b], so we used these preferentially over other data available on GenBank that might include misidentifications or different taxon concepts.
However, using this dataset alone for phylogenetic analysis suffers two drawbacks: it is not densely sampled enough to resolve some internal nodes, and it lacks many lineages that have fossils available for molecular dating.
Furthermore, use of community phylogenies generated by only by sampling the species present in the local community has been shown to produce spurious results in simulation studies [@Park2018].
Because one goal of the study is to understand the distribution of paleo- vs. neoendemic areas (defined in units of time), we require an ultrametric tree.
Therefore, to obtain a robust, ultrametric tree, we combined the *rbcL* data of @Ebihara2019b (`r n_japan_fern_taxa` taxa) with a globally sampled plastid dataset including all fern sequences on GenBank for four widely sequenced plastid genes (*atpA*, *atpB*, *rbcL*, and *rps4*; `r n_ftol_fern_taxa` taxa) and `r n_plastome_genes` other coding, single-copy plastid genes extracted from `r n_plastome_fern_taxa` complete fern plastomes [FTOL v0.0.1\; @Nitta2022b] (`r ape::Ntip(plastome_tree) %>% scales::number(big.mark = ",")` operational taxonomic units [OTUs] total, including outgroup taxa).
This gene sampling is comparable to a recent global fern phylogeny that resolved relationships across ca. 4,000 taxa using six plastid markers [@Testo2016a], and the addition of plastome data can be expected to increase support along the backbone.

We conducted maximum-likelihood phylogenetic analysis with `r software("IQ-TREE")` v1.6.12 [@Nguyen2015] under the GTR+I+G model (all sites concatenated into a single data matrix), and assessed node support with 1,000 ultrafast bootstrap replicates [@Minh2013].
Molecular dating was conducted with the program `r software("treePL")` v1.0 [@Smith2012] using `r n_fossil_points` fossil calibration points after @Testo2016a, with the exception of treating fossil *Kuylisporites* as belonging to crown *Cyathea*, not *Cyathea* + *Alsophila* [@Loiseau2020].
The root of the tree (most recent common ancestor of all land plants) was fixed to 475 mya [@Testo2016a].
We trimmed the resulting ultrametric tree to only Japanese taxa.
`r software("treePL")` requires branch lengths to be >0, and it sets extremely small branches to an arbitrary minimum positive length.
This behavior may artificially add branch length between taxa with identical DNA sequences, which occur multiple times in the Japanese fern *rbcL* dataset.
We converted any clades in the dated tree consisting of identical sequences to polytomies with the node at 0 million years ago (i.e., the present; Ma).
This tree was used for all subsequent analysis unless indicated otherwise.

@@

We tested for phylogenetic signal in continuous traits using two alternative metrics, Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999].
*K* describes the ratio between the observed variance in a trait vs. the amount of variance expected under brownian motion (BM); $\lambda$ is a scaling parameter between 0 and 1 that transforms the phylogeny such that trait values fit those most closely expected under BM.
Both $\lambda$ and *K* = 1 under BM; relatively lower values indicate less signal (overdispersion) and higher values indicate more signal (clustering).
We measured *K* and $\lambda$ using the `r func("phylosig")` function in the R package `r pack("phytools")` [@Revell2012], which also calculates significance based on a randomization test for *K* and a likelihood ratio test for $\lambda$.
We tested for phylogenetic signal in qualitative (coded as binary) traits using Fritz and Purvis' *D* [@Fritz2010] with the `r func("phylo.d")` function in the R package `r pack("caper")` [@Orme2018], which also assesses significance by comparing the observed value to a null distribution of random values and a simulated distribution of values expected under BM.
Values of *D* range from 0 under BM to 1 under random evolution, but can exceed this range in cases of extreme clumping or overdispersion, respectively.
These methods for phylogenetic signal cannot use a tree with zero-length polytomies, so we used the original dated tree from treePL.

@@

We measured taxonomic diversity using taxonomic richness (`r ie`, number of taxa in each community).

@@

We measured functional diversity (FD) using the method of @Petchey2002.
We first calculated functional distances using Gower's method [@Gower1971] on the trait data with the `r func("gowdis")` function in the R package `r pack("FD")` [@Laliberte2010].
We weighted continuous and qualitative traits equally, weighted each binary component trait of each qualitative trait equally, and log-transformed and scaled continuous traits prior to calculating distances.
We also tested alternative weighting schemes, but these produced very similar results (data not shown).
We then used the distances to build a trait dendrogram using the `r func("hclust")` function in the R package `r pack("stats")` [@RCoreTeam2021], and used the dendogram in place of a phylogenetic tree to calculate PD and RPD as described above.
These functional analogs of PD and RPD are hereafter referred to as FD (functional diversity) and RFD (relative functional diversity).

@@

Observed values of PD, RPD, FD, RFD, and PE are known to be related to species richness; as more taxa are drawn at random without replacement from a given set, it becomes increasingly unlikely to draw a taxon that is distantly related (or functionally dissimilar) to the others.
Therefore, we determined whether the observed metrics are more or less diverse than random for a given number of taxa by conducting a randomization test.
We generated 999 random communities using the curveball algorithm of @Strona2014, which conserves richness per site and species occurrence frequencies while randomizing species identities, then compared the observed value to the distribution of random values.
Statistical significance was assessed using a one-tailed test for PE or a two-tailed test for other metrics with $\alpha$ = 0.5; observed values in the extreme 5% of the null distribution were considered significant.
We also calculated the standard effect size (SES) of each biodiversity metric.
SES measures how extreme the observed value (Obs) is relative to the null distribution (Null):

$SES = \frac{Obs - Mean(null)}{SD(Null)}$

We tested whether centers of endemism have an over-representation of old (paleoendemic) or new (neoendemic) lineages using categorical analysis of neo- and paleoendemism (CANAPE), which involves measuring PE with an alternate tree where all branch lengths are equal, then comparing the ratio between raw PE and alternative PE (RPE) [@Mishler2014].
Low RPE indicates short range-restricted branches (neoendemism) and high RPE indicates long range-restricted branches (paleoendemism).
Since all of these measures are affected by species richness, categorization of endemism is done using the rank of each observed variable relative to those calculated for the null distribution. All measures of biodiversity and CANAPE were carried out using the R package `r pack("canaper")` [@Nitta2022].
